Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions

### 2022-03-23 analysis/paper/shceof.Rmd
Revisión

```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)", fig.height=1.2*width_column, fig.width=width_column, dependson="plot_regr"}
plot_regr(regr , "cEOF1", c(200, 50), "hgt") +
  scale_fill_divergent(mgp,
                       breaks = AnchorBreaks(0, exclude = 0),
                       guide = guide_colorsteps_bottom(15, even.steps = TRUE), 
                       limits = c(-1800, 1800)) 
```

Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º out of phase.
Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over East Antarctica.
Instead of a wave-1 pattern, regressed anomalies describe a monopole with its centre displaced from the South Pole.
The latter indicates an influence of cEOF1 in modulating the polar vortex.

The regression maps for cEOF1 at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
It is only possible to partially distinguish the wave 1 pattern in relation with the Real cEOF1 (Figure \@ref(fig:eof1-regr-gh)b.1).
The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.

With the exception of the Imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig. \@ref(fig:ceofs-1)), particularly at 200 hPa.
Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º, suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
While in the troposphere, they are associated with slightly off-centre monopoles.

(ref:eof2-regr-gh-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second cEOF.

@@

Figure \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height anomalies upon the cEOF2.
Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure \@ref(fig:ceofs-1).
Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans.
cEOF2 then represents an equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001].
Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the Real cEOF2 can roughly be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.

```{r zonal-waves, fig.cap = "Amplitud de ondas zonales de la regresión con altura geopotencial (figuras anteriores)"}
# regr %>%
#   .[lat <= -20] %>%
#   .[lev %in% c(200, 50)] %>%
#   # .[cEOF == "cEOF1"] %>%
#   .[, FitWave(estimate, 0:3), by = .(lev, cEOF, term, lat)] %>%
#   # .[, pseudor2 := amplitudepseudor2^2/sum(amplitude^2), by = .(lev, cEOF, term, lat)] %>% 
#   ggplot(aes(lat, amplitude)) +
#   geom_line(aes(color = factor(k))) +
#   ggh4x::facet_nested(cEOF + lev ~ term, scales = "free", labeller = labeller(lev = lab_lev)) +
#   scale_color_brewer("Wavenumber", palette = "Dark2") +
#   scale_y_continuous("Amplitude") +
#   scale_x_latitude() +
#   coord_flip() +
#   panel_background
```

>>>>

```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)", fig.height=1.2*width_column, fig.width=width_column, dependson="plot_regr"}
plot_regr(regr , "cEOF1", c(200, 50), "hgt") +
  scale_fill_divergent(mgp,
                       breaks = AnchorBreaks(0, exclude = 0),
                       guide = guide_colorsteps_bottom(15, even.steps = TRUE), 
                       limits = c(-2000, 2000)) 
```

Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), the Real cEOF1 is associated with a monopole centred over the Ross Sea.
The Imaginary cEOF1 is associated with a clear wave 1 pattern with maximum over the coast of East Antarctica.

At 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) the Real cEOF1 shows a single area of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes with is centre shifted slightly eastward compared with the upper level anomalies.
The Imaginary cEOF2.
The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase [@fogt2020].
Both components of cEOF1 could be associated to zonal anomalies in the upper troposphere jet stream intensity.

Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
While in the troposphere, they are associated with slightly off-centre monopoles.

(ref:eof2-regr-gh-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for cEOF2.

@@

Figure \@ref(fig:eof2-regr-gh) shows the regression maps of geopotential height anomalies upon the cEOF2.
Wave trains similar to those identified for cEOF2 patterns (Fig \@ref(fig:ceofs-1)) can be distinguished.
Regressed anomalies associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
All fields have a dominant zonal wave 3 limited to the western hemisphere, over the Pacific and Atlantic Oceans.
cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns [@mo2001].
Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the Real cEOF2 regression map could be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.
